Tidal streams in triaxial systems

### 2015-02-14 ms.tex
intro noise

@@

The halos of galaxies are filled with substructure...
As a satellite galaxy or globular cluster orbits within a parent galaxy, its mass is eroded due to the tidal forces of the host galaxy. For (typically) many orbits, these tidally stripped stars remain spatially coherent.
Of particular interest are cold, young dynamical structures --- ``tidal streams'' -- which provide useful information about the mass distributions at large distances in the galaxies within which they orbit. [At these distances, disk / classic baryonic components can't help you, mostly dark matter]
The formation and morphology of tidal streams has been studied extensively in a variety of assumed potential forms. [Stream formation in oblate, spherical potentials, action-angle, etc.]

\hrule

For an integrable Hamiltonian, $H$, with $N$ degrees of freedom (dof), the equations of motion are most simply expressed in angle-action coordinates. In these coordinates, the actions, $J_n$, are integrals of the motion, and the conjugate angle variables, $\theta_n$, increase linearly with time
\begin{align}
	\act_n &= {\rm const.}\\
	\dot{\act}_n &= -\frac{\partial H}{\partial \theta_n} = 0\\
	\theta_n(t) &= \Omega_n t + \theta_n(0)\\
	\dot{\theta}_n &= \frac{\partial H}{\partial \act_n} = \Omega_n(\act_1,...,\act_n)\label{eq:aafreq}
\end{align}
where $i=1,2,...N$, and $\Omega_n$ are the fundamental frequencies of the orbit. Owing to the involutive property of the actions and the periodic nature of the angle variables, the topology of angle-action space is equivalent to that of the surface of an $N$-dimensional torus, and thus orbits are often referred to in terms of \emph{orbital tori}: each set of actions, $(\act_1,\act_2,...\act_N)$, uniquely determine the torus associated with a given orbit and the angles specify the location on the surface of the $N$-torus.

@@

Bounded, regular orbits in a triaxial potential will have three fundamental frequencies, $\bs{\Omega}$, that determine the periodic behavior of motion (Eq.~\ref{eq:aafreq}). This implies that for any canonical coordinate, $q$, the motion in this coordinate can be decomposed as 
\begin{equation}
	q(t) = \sum^\infty_k a_k e^{i \omega_k t}
\end{equation}
where these frequencies, $\omega_k$, are linear, integer combinations of the fundamental frequencies
\begin{equation}
	\omega_k = \bs{n}_k \cdot \bs{\Omega}
\end{equation}
and the $a_k$ are complex amplitudes. \cite{laskar93} introduced a method for recovering the fundamental frequencies of an orbit that effectively uses a Hanning-filtered fast-Fourier transform (FFT) of complex combinations of the motion (e.g., $x(t) + i v_x(t)$). It has been shown that using the Hanning filter makes the accuracy in determining the frequencies scale as $\inttime^{-4}$ \citep{laskar99}, where $\inttime$ is the integration time. We follow \cite{valluri98} and refer to this method as ``Numerical Approximation of Fundamental Frequencies'' (NAFF) even though it differs slightly from the original method from Laskar.

@@

\subsection{Orbital properties: frequencies and frequency diffusion}

@@

For each orbit, we also compute the frequency diffusion rate per orbit, $\fdrate$ (Equation~\ref{eq:fdrate}) by computing the frequencies for the first half and second half of the orbital time series separately. This rate is only a measure of the time-averaged diffusion rate because the frequencies are computed for long segments of orbit, but \cite{valluri??} [...]. Figure~\ref{fig:diffusionmap} shows maps of the frequency diffusion for the $xz$ (left) and $yz$ (right) grids: plotted are the initial conditions for all orbits colored by $\log_{10}(\fdrate)$ (white points are strongly chaotic, black are regular). The rich structure of unstable resonances apparent in these frequency diffusion maps highlight the accuracy of NAFF. It seems that tube orbits in these potentials are mostly regular or only mildly chaotic, however islands of stronger chaos do appear, especially at the intersections of unstable resonances. The bright, unstable regions in these maps are a visualization of the Arnold web at this energy. It is evident that these unstable regions are interconnected and therefore a tube orbit that starts, for example, in the lower left-hand region of one of these maps could  theoretically end up as a box orbit (along the zero-velocity curve) at some point in it its history (but the timescale for this to happen is likely prohibitively long).

@@

@@

\subsection{Ensemble properties: mixing of nearby orbits}

Look at minimum value of $D_{\rm KL}$ over $T_{\rm ev}$, use that as metric of

@@

@@

\acknowledgements
The authors wish to thank ...
APW is supported by a National Science Foundation Graduate Research Fellowship under Grant No.\ 11-44155. 
This research made use of Astropy, a community-developed core \texttt{Python} package for Astronomy \citep{astropy13}.
This work additionally relied on Columbia University's \emph{Hotfoot} and \emph{Yeti} compute clusters, and we acknowledge the Columbia HPC support staff for assistance.

>>>>

On the largest scales ($\gtrsim$30~kpc for a Milky-Way-like galaxy) in the halos of galaxies, the gravitational potentials are thought to be triaxial. However, despite suggestive evidence, this fundamental prediction from LCDM has not been conclusively confirmed from observations. Around other galaxies, it is generally hard to measure the 3D shape of [..] because they are seen in projection. From our position within the Milky Way, our 3D view of our own halo and proximity gives us a unique chance to directly measure the 6D positions of stars and model the shape of the mass distribution at large radii. Unfortunately, the halo also has a low density of tracers. Luckily, however, many of the halo stars are likely associated with debris stripped from in-falling satellite galaxies.

As a satellite galaxy or globular cluster orbits within some larger system, mass is eroded due to the tidal forces of the host galaxy potential. The phase-space density and therefore the observable properties of the tidal debris are sensitive to the internal progenitor properties, orbit of the progenitor, and underlying mass distribution of the host galaxy. The morphological evolution of the debris depends on the spread of orbital properties (e.g., actions) of the debris and the orbit of the progenitor system, both of which are determined by the shape and radial profile of the gravitational potential of the host galaxy. In the halo of the Milky Way, tidal streams of stars are found over a large range of distances. By simultaneously modeling the observed density of stream stars along with the potential, it is hoped that we may infer the dark matter distribution of the Milky Way on these large scales.

Tidal streams are generally modeled as 1D structures with simple internal progenitor dynamics and density evolution. 
Pearson et al. (2014) tried to reproduce observations of the stellar stream density from the globular cluster Palomar 5 in an oblate and a triaxial potential using \emph{Streakline} models (K\"upper et al. (XXXX)) : 
		\item in the oblate potential, a thin model stream is easily found that reproduces the observed stellar density morphology of the stream;
		\item in the triaxial potential, the model streams generically form large, two-dimensional ``fans'' of debris near the ends, and there are no physically reasonable progenitor orbits that reproduce the observed thinness of the stream, given the observational constraints of the present-day position and velocity of the cluster.
This result suggests that the shape of the host galaxy potential directly influences the morphology of the resulting debris, and that other morphological classes exist between thin, 1D streams and diffuse, 2D shells.
To understand the formation of these fans and the implications of this work we must first understand the complex orbital structure of triaxial systems.

In this work, we study the effect of orbital classes that are prevalent in triaxial systems (e.g., stochastic, near-resonant, resonant, and quasiperiodic) on the resulting morphology of tidal streams.

For (typically) many orbits, these tidally stripped stars remain spatially coherent.
Of particular interest are cold, young dynamical structures --- ``tidal streams'' -- which provide useful information about the mass distributions at large distances in the galaxies within which they orbit. [At these distances, disk / classic baryonic components can't help you, mostly dark matter]
The formation and morphology of tidal streams has been studied extensively in a variety of assumed potential forms. [Stream formation in oblate, spherical potentials, action-angle, etc.]

@@

The halos of galaxies are filled with substructure...

For an integrable Hamiltonian, $H$, with $N$ degrees of freedom (dof), the equations of motion are most simply expressed in angle-action coordinates. In these coordinates, the actions, $J_n$, are integrals of the motion, and the conjugate angle variables, $\theta_n$, increase linearly with time
\begin{align}
	\act_n &= {\rm const.}\\
	\dot{\act}_n &= -\frac{\partial H}{\partial \theta_n} = 0\\
	\theta_n(t) &= \Omega_n t + \theta_n(0)\\
	\dot{\theta}_n &= \frac{\partial H}{\partial \act_n} = \Omega_n(\act_1,\act_2,...,\act_n)\label{eq:aafreq}
\end{align}
where $i=1,2,...N$, and $\Omega_n$ are the fundamental frequencies of the orbit. Owing to the involutive property of the actions and the periodic nature of the angle variables, the topology of angle-action space is equivalent to that of the surface of an $N$-dimensional torus, and thus orbits are often referred to in terms of \emph{orbital tori}: each set of actions, $(\act_1,\act_2,...\act_N)$, uniquely determine the torus associated with a given orbit and the angles specify the location on the surface of the $N$-torus.

@@

Bounded, regular orbits in a triaxial potential will have three fundamental frequencies, $\bs{\Omega}$, that determine the periodic behavior of motion (Eq.~\ref{eq:aafreq}). This implies that for any canonical coordinate, $q$, the motion in this coordinate can be decomposed as 
\begin{equation}
	q(t) = \sum^\infty_k a_k e^{i \omega_k t}
\end{equation}
where these frequencies, $\omega_k$, are linear, integer combinations of the fundamental frequencies
\begin{equation}
	\omega_k = \bs{n}_k \cdot \bs{\Omega}
\end{equation}
and the $a_k$ are complex amplitudes and $\bs{n}_k$ is an integer vector. \cite{laskar93} introduced a method for recovering the fundamental frequencies of an orbit that effectively uses a Hanning-filtered fast-Fourier transform (FFT) of complex combinations of the motion (e.g., $x(t) + i v_x(t)$). It has been shown that using the Hanning filter makes the accuracy in determining the frequencies scale as $\inttime^{-4}$ \citep{laskar99}, where $\inttime$ is the integration time. We follow \cite{valluri98} and refer to this method as ``Numerical Approximation of Fundamental Frequencies'' (NAFF) even though it differs slightly from the original method from Laskar.

@@

\subsection{Orbital properties: frequencies, frequency diffusion, and Lyapunov exponents}

@@

For each orbit, we also compute the frequency diffusion rate per orbit, $\fdrate$ (Equation~\ref{eq:fdrate}) by computing the frequencies for the first half and second half of the orbital time series separately. This rate is only a measure of the time-averaged diffusion rate because the frequencies are computed for long segments of orbit, but \cite{valluri??} [...]. Figure~\ref{fig:diffusionmap} shows maps of the frequency diffusion for the $xz$ (left) and $yz$ (right) grids: plotted are the initial conditions for all orbits colored by $\log_{10}(\fdrate)$ (white points are strongly chaotic, black are regular). The rich structure of unstable resonances apparent in these frequency diffusion maps highlight the accuracy of NAFF. It seems that tube orbits in these potentials are mostly regular or only mildly chaotic, however islands of stronger chaos do appear, especially at the intersections of unstable resonances. The bright, unstable regions in these maps are a visualization of the Arnold web at this energy. It is evident that these unstable regions are interconnected and therefore a tube orbit that starts, for example, in the lower left-hand region of one of these maps could  theoretically end up as a box orbit (along the zero-velocity curve) at some point in it its history (but the timescale for this to happen is likely prohibitively long). Figure~\ref{fig:diffusionrates} shows histograms of $\log_{10}(\fdrate)$ for all orbits in each initial condition grid for box and tube orbits separately. [...1-2 sentences discussion this (need to wait to see final figure)]

@@

\begin{figure}[h]
\centering
	\includegraphics[width=0.45\columnwidth] {figures/diffusion_rate_hist_xz.pdf}
	\includegraphics[width=0.45\columnwidth] {figures/diffusion_rate_hist_yz.pdf}
	\caption{Not the final figures. Need to re-make with same axis limits, remove labels, put in same PDF file...etc. Separate box from tube orbits.} 
	\label{fig:diffusionrates}
\end{figure}

[Lyapunov exponent grids...]

@@

\subsection{Ensemble properties: mixing of debris}

We measure the state of mixing of the debris as described above: we compute the KL divergence between the configuration-space density of the ensemble and the configuration-space density expected if the ensemble particles were uniformly distributed over the parent orbit energy hypersurface.

a:
This measure can be computed at any time -- [we show two different maps of this quantity: one after 10 Gyr, one after 15 orbits?]

b: 
From the KL divergence, we compute a mixing time by measuring the slope or power law index?

[Measure mixing time from KLD(t), plot freq. diffusion time vs. mixing time, show we can predict morphology from orbital properties of progenitor?]

\subsection{Observable consequences of mixing}

Only effectively looking at debris disrupted from a single pericenter -- debris near progenitor constantly repopulated as more and more stuff disrupts out

Streams should be shorter? How to test this? Length of stream in angle at isodensity contour?

@@

Recovering the potential of triaxial systems is not trivial. Danger of orbit fitting because chaotic orbits can reproduce morphologies of regular orbits.

@@

\acknowledgements
The authors wish to thank Sarah Pearson, Andreas K\"upper, David Hogg, David Merritt, \emph{Stream Team}, etc...
APW is supported by a National Science Foundation Graduate Research Fellowship under Grant No.\ 11-44155. 
This research made use of Astropy, a community-developed core \texttt{Python} package for Astronomy \citep{astropy13}.
This work additionally relied on Columbia University's \emph{Hotfoot} and \emph{Yeti} compute clusters, and we acknowledge the Columbia HPC support staff for assistance.
